---
UOLID: "220460044"
Subject Code: "ST2195"
---
# Part (1)

## (a)

### Given probability density function
```{r}
f <- function(x) {
  return(0.5 * exp(-abs(x)))# defining f(x)
}
```

### Step 1:
```{r}
x0 <- 0     # initial x value
N <- 10000  # sample size
s <- 1      # standard deviation
```


### Step 2: Function to perform Metropolis-Hastings Algorithm.
```{r}
# setting up Metropolis-Hasting Algorithm
mh <- function (N, x0, s){
  x <- numeric(N) # setting up numeric vector to store generated x
  x_current <- x0 
  
  for (i in 1:N) {
    x_star <- rnorm(1, mean = x_current, sd = s)  # simulate random x* from Normal distribution
    
    # computing log acceptance ratio
    lg_r <- log(f(x_star)) - log(f(x_current)) # log is taken to make sure linearity
    
    # generate random number u
    lg_u <- log(runif(1)) # random u from uniform distribution from [0,1)
    
    # accept or reject x0 based on log acceptance ratio
    if (lg_u < lg_r) {
      x_current <- x_star # accepted
    } else {
      x_current <- x_current # rejected
    }
    
    # saving the sample
    x[i] <- x_current
  }
  
  return(x)
}
```

### Running Metropolis-Hastings Algorithm.
```{r}
x <- mh(N, x0, s)

# To plot a histogram
hist(x, main = "Metropolis-Hastings Samples", xlab = "X", prob = TRUE)

# To plot a kernel density plot
lines(density(x), col = 'purple',lwd = 1)

# Plotting f(x) distribution into existing plot
curve(f(x), from = min(x), to = max(x), col = 'pink', add = TRUE, lwd = 1)
```

### Reporting Sample mean and Standard deviation of generated samples.
``` {r}
montecarlo_mean <- mean(x)
montecarlo_sd <- sd(x)

# Printing results
cat ("Monte Carlo Estimate of Mean:", montecarlo_mean, "\n") # cat is used to concatenate the values into a single string
cat ("Monte Carlo Estimate of Standard Deviation:", montecarlo_sd, "\n")
```

## (b)

## Part b parameters.
```{r}
N_b <- 2000
s_b <- 0.001
J_b <- 4
```

### Obtaining chain sequence.
```{r}
mh_chains <- function (N_b, s_b, J_b){
  chains <- matrix (NA, nrow = N_b, ncol = J_b) # matrix to store chain, all values of (n,j) is NA

for (j in 1:J_b){
  x_current_b <- 0 # Setting all initial value of each chain as 0
  
  for (i_b in 1:N_b){
    x_star_b <- rnorm(1, mean = x_current_b, sd = s_b) #simulate X*~N (x_current, s^2)
    
    # computing log acceptance ratio
    lg_r_b <- log(f(x_star_b)) - log(f(x_current_b)) # log is taken to make sure linearity
    
    # generate random number u
    lg_u_b <- log(runif(1)) # random u from uniform distribution from [0,1)
    
    # accept or reject x0 based on log acceptance ratio
    if (lg_u_b < lg_r_b) {
      x_current_b <- x_star_b
    }
    
    chains[i_b,j] <- x_current_b # i and j defined in line 96 and 93 respectively
  }
}

return(chains)
}
```


## Computation:

```{r}
# generating chain with parameters
chains_b <- mh_chains (N_b, s_b, J_b) 

# checking for NA in chain
if(any(is.na(chains_b))){
  warning ("There in NA in chain_b")
}
```

### Sample mean of chain j, Mj
```{r}
# function to loop through each chain and calculate sample mean
sample_mean_chain <- function (chains_b, J_b){
  M_j <- numeric (J_b)

  for (j in 1:J_b) {
    chain_j <- chains_b [,j]
    M_j[j] <- mean(chain_j, na.rm = TRUE)
  }
  return (M_j)
}

# calculating sample mean for each chain
M_j <- sample_mean_chain(chains_b, J_b)

# print
for (j in 1:J_b){
  cat ("The sample mean, Mj, of chain j =", j, ":", M_j[j], "\n")
}
```

### Sample variance of chain j, Vj
```{r}
# function for sample variance of chain j (looped)
sample_variance_chain <- function (chains_b, J_b) {
  V_j <- numeric (J_b)
  
  for (j in 1:J_b){
    chain_j <- chains_b [,j]
    V_j[j] <- sum((chain_j - mean(chain_j, na.rm = NA))^2)/N_b
  }
  return (V_j)
}

# calculating sample variance for each chain
V_j <- sample_variance_chain (chains_b, J_b)

# print
for (j in 1:J_b){
  cat ("The sample variance, Vj, of chain j =", j, ":", V_j[j], "\n")
}
```
### Overall within sample variance, W
```{r}
# function for overall within sample variance W
overall_within_variance <- function (chains_b, J_b){
  W <- numeric (J_b)

  W <- sum(V_j)/J_b 
  
  return (W)
}

# calculating overall within sample variance W
W <- overall_within_variance(chains_b, J_b)

# print
cat ("The overall within sample variance, W:", W, "\n")

```

### Overall sample mean, M
```{r}
# function for overall sample mean M
overall_sample_mean <- function (chains_b, J_b) {
  M <- numeric (J_b)
  
  M <- sum(M_j)/J_b 
  
  return (M)
}

# calculating overall sample mean M
M <- overall_sample_mean(chains_b, J_b)

# print
cat ("The overall sample mean, M:", M,  "\n")
```

### In between sample variance B
```{r}
# function for between sample variance B
inbtw_var_B <- function (chains_b, J_b) {
  B <- numeric (J_b)

  B <- sum(M_j - M)^2 / J_b
  
  return (B)
}

# calculating between sample variance variance B
B <- inbtw_var_B (chains_b, J_b)

# print
cat ("The between sample variance, B:", B,  "\n")

```

### R hat value
```{r}
# function for r hat
calc_r_hat <- function (chains_b, J_b) {
  r_hat <- numeric (J_b)
  
  r_hat <- sqrt ((B+W)/ W)

  
  return (r_hat)
}

# calculating r hat
r_hat <- calc_r_hat (chains_b, J_b)

# print
cat ("The r hat value is", r_hat, "\n")

```

### Keeping N and J fixed, plot r hat over s [0.001, 1]
```{r}
# set seed to ensure reproducibility
set.seed(123)

# Metropolis-Hastings algorithm to generate chains for a given s
metropolis_hastings <- function(x0, N_b, s, j) {
  samples <- numeric(N_b) #Creating an empty vector with length N_b
  current_value <- x0
  
  for (i in 1:N_b) {
    x_star <- rnorm(1, mean = current_value, sd = s)
    log_ratio <- -abs(x_star)^j + abs(x0)^j  # Old ratio with logarithm
    log_u <- log(runif(1))
    
    if (log_u < log_ratio) {
      current_value <- x_star
    }
    
    samples[i] <- current_value
  }
  
  return(samples)
}

# Function to calculate R_hat for a given s
calc_R_hat <- function(N, s, J, j) {
  chains <- replicate(J, metropolis_hastings(0, N, s, j))
  M_j <- colMeans(chains, na.rm = TRUE)
  V_j <- apply(chains, 2, var, na.rm = TRUE)
  W <- mean(V_j)
  M <- mean(M_j)
  B <- mean((M_j - M)^2)
  
  # Ensure B is not zero to avoid division by zero
  if (B == 0) {
    B <- 1e-6  # Small nonzero value
  }
  
  R_hat <- sqrt((B + W) / W)
  
  return(R_hat)
}

# Parameters
N_s <- 2000
J_s <- 4
s_values <- seq(0.001, 1, length.out = 100)
j_s <- 2  # used for raising power to 2

# Calculate R_hat for each s
R_hat_values <- sapply(s_values, calc_R_hat, N = N_s, J = J_s, j = j_s)

# Plot the results
plot(s_values, R_hat_values, type = "l", xlab = "s", ylab = "R_hat", main = "R_hat against s values")
```